Carrier phase and amplitude estimation using pilots and data

### 2012-03-28 paper.tex
Added IEEEbst and fixed other minor stuff.
added ant build script.
Added some refs

In this section we derive Mackentun's~\cite{Mackenthun1994} least squares estimator for the coherent case.  Mackenthun specifically consider the noncoherent case, but only minor modifications are required to meaninfully include the pilot symbols, and the fundamental ideas in this section come from~\cite{Mackenthun1994}.  For the purpose of anaylsing computational complexity, we will assume that the number of data symbols $\abs{D}$ is proportional to the total number of symbols $L$, so that, for example, $O(L) = O(\abs{D})$.  In this case Mackentun's algorithm requires $O(L \log L)$ arithmetic operations.  This complexity arises from the need to sort a list of $\abs{D}$ elements.  In Section~\ref{sec:line-time-algor} we will show that a full sort is not neccesary, and that the least squares estimator can be implemented in $O(L)$ operations.

Define the sum of squares function
\begin{align}
SS(a, \{d_i, i \in D\}) &= \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \nonumber \\
&= \sum_{i \in P \cup D} \abs{y_i}^2 - a s_i y_i^* - a^* s_i^* y_i + aa^*. \label{eq:SS}
\end{align}
Fixing the $\{d_i, i \in D\}$, differentiating with respect to $a^*$, and setting the resulting equation to zero we find that the least squares estimator of $a_0$, satisfies
\begin{equation}\label{eq:hata}
\hat{a} = \frac{1}{L} \sum_{i \in P \cup D} y_i s_i^* = \frac{Y}{L}
\end{equation}
where $L = \abs{P \cup D}$ is the total number of symbols transmitted, and to simplify our notation we put 
\[
Y = \sum_{i \in P \cup D} y_i s_i^* = \sum_{i \in P } y_i p_i^* + \sum_{i \in D } y_i d_i^*
\]  
Note that $Y$ is a function of the unknown data symbols $\{ d_i, i \in D\}$ and we could write $Y(\{ d_i, i \in D\})$, but have chosen to surpress the argument $(\{ d_i, i \in D\})$ for notational clarity.  Substituting $\hat{a} = \frac{Y}{L}$ into~\eqref{eq:SS} we obtain the sum of squares function, conditioned on minimisation with respect to $a$,
\begin{equation}\label{eq:SSdatasymbols}
SS(\{d_i, i \in D\}) = A - \frac{1}{L}\abs{Y}^2,
\end{equation}
where $A = \sum_{i \in P \cup D}\abs{y_i}^2$ is a constant.  Observe that given a candidate value for the data symbols, we can compute the corresponding sum of squares function $SS(\{d_i, i \in D\})$ in $O(L)$ arithmetic operations.  We will show that there are atmost $(M+1)\abs{D}$ candidate values of the least squares estimator of the data symbols~\cite{Sweldens2001,Mackenthun1994}.  %When the number of data symbols is not small, this set is substantially smaller than the entire set of possible transmitted symbols, which is of size $M^{\abs{D}}$.

To see this, let $a = \rho e^{j\theta}$.  Now the sum of squares function can be written as
\begin{align*}
SS(\rho, \theta, \{d_i, i \in D\}) &= \sum_{i \in P \cup D} \abs{ y_i - \rho e^{j\theta} s_i }^2  \\
&= \sum_{i \in P} \abs{ y_i - \rho e^{j\theta} p_i }^2 + \sum_{i \in D} \abs{ y_i - \rho e^{j\theta} d_i }^2.
\end{align*}
For given $\theta$, the least squares estimator of the $i$th data symbol $d_i$ is given by minimising $\abs{ y_i - \rho e^{j\theta} d_i }^2$, that is,
\begin{equation}\label{eq:hatdfinxtheta}
\hat{d}_i(\theta) = \exp\left( j\frac{2\pi}{M}\hat{u}_i(\theta) \right) \qquad \text{where} \qquad \hat{u}_i(\theta) = \round{\frac{M}{2\pi}\angle( e^{-j\theta}y_i)},
\end{equation}
where $\round{\cdot}$ rounds its argument to the nearest integer\footnote{The direction of rounding for half-integers is not important so long as it is consistent.  The authors have chosen to round up half-integers in their own implementation.}.  Note that $\hat{d}_i(\theta)$ does not depend on the amplitude $\rho$.  As defined, $\hat{u}_i(\theta)$ is not strictly inside the set of integers $\{0, 1, \dots, M-1\}$, but this is not of consequence, as we intend its value to be considered modulo $M$.  With this in mind let $\phi_i = \angle y_i$, then,
\[
\hat{u}_i(\theta) = \round{\frac{M}{2\pi}( \phi_i - \theta )}
\]
which is equivalent to the definition from~\eqref{eq:hatdfinxtheta} modulo $M$.

@@

@@

\begin{figure*}[tp]
	\centering
		\includegraphics{code/plot-2.mps}
		\caption{Phase error}
		\label{fig:plotamp}
\end{figure*}

\begin{figure*}[tp]
	\centering
		\includegraphics{code/plot-3.mps}
		\caption{Complex gain error}
		\label{fig:plotamp}
\end{figure*}

@@

\bibliography{../../bib/bib}

>>>>

In this section we derive Mackentun's~\cite{Mackenthun1994} least squares estimator.  Mackenthun specifically considered the noncoherent setting, but only minor modifications are required to meaninfully include the pilot symbols, so we also consider the coherent setting here.  For the purpose of anaylsing computational complexity, we will assume that the number of data symbols $\abs{D}$ is proportional to the total number of symbols $L$, so that, for example, $O(L) = O(\abs{D})$.  In this case Mackentun's algorithm requires $O(L \log L)$ arithmetic operations.  This complexity arises from the need to sort a list of $\abs{D}$ elements.  In Section~\ref{sec:line-time-algor} we will show that a full sort is not neccesary, and that the least squares estimator can be implemented in $O(L)$ operations.

Define the sum of squares function
\begin{align}
SS(a, \{d_i, i \in D\}) &= \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \nonumber \\
&= \sum_{i \in P \cup D} \abs{y_i}^2 - a s_i y_i^* - a^* s_i^* y_i + aa^*. \label{eq:SS}
\end{align}
Fixing the data symbols $\{d_i, i \in D\}$, differentiating with respect to $a^*$, and setting the resulting equation to zero we find that the least squares estimator of $a_0$, satisfies
\begin{equation}\label{eq:hata}
\hat{a} = \frac{1}{L} \sum_{i \in P \cup D} y_i s_i^* = \frac{1}{L} Y
\end{equation}
where $L = \abs{P \cup D}$ is the total number of symbols transmitted, and to simplify our notation we put 
\[
Y = \sum_{i \in P \cup D} y_i s_i^* = \sum_{i \in P } y_i p_i^* + \sum_{i \in D } y_i d_i^*
\]  
Note that $Y$ is a function of the unknown data symbols $\{ d_i, i \in D\}$ and we could write $Y(\{ d_i, i \in D\})$, but have chosen to surpress the argument $(\{ d_i, i \in D\})$ for notational clarity.  Substituting $\hat{a} = \frac{1}{L}Y$ into~\eqref{eq:SS} we obtain the sum of squares function, conditioned on minimisation with respect to $a$,
\begin{equation}\label{eq:SSdatasymbols}
SS(\{d_i, i \in D\}) = A - \frac{1}{L}\abs{Y}^2,
\end{equation}
where $A = \sum_{i \in P \cup D}\abs{y_i}^2$ is a constant.  Observe that given a candidate value for the data symbols, we can compute the corresponding sum of squares function $SS(\{d_i, i \in D\})$ in $O(L)$ arithmetic operations.  We will show that there are atmost $(M+1)\abs{D}$ candidate values of the least squares estimator of the data symbols~\cite{Sweldens2001,Mackenthun1994}.  %When the number of data symbols is not small, this set is substantially smaller than the entire set of possible transmitted symbols, which is of size $M^{\abs{D}}$.

To see this, let $a = \rho e^{j\theta}$.  Now the sum of squares function can be written as
\begin{align*}
SS(\rho, \theta, \{d_i, i \in D\}) &= \sum_{i \in P \cup D} \abs{ y_i - \rho e^{j\theta} s_i }^2  \\
&= \sum_{i \in P} \abs{ y_i - \rho e^{j\theta} p_i }^2 + \sum_{i \in D} \abs{ y_i - \rho e^{j\theta} d_i }^2.
\end{align*}
For given $\theta$, the least squares estimator of the $i$th data symbol $d_i$ is given by minimising $\abs{ y_i - \rho e^{j\theta} d_i }^2$, that is,
\begin{equation}\label{eq:hatdfinxtheta}
\hat{d}_i(\theta) = \exp\left( j\frac{2\pi}{M}\hat{u}_i(\theta) \right) \qquad \text{where} \qquad \hat{u}_i(\theta) = \round{\frac{M}{2\pi}\angle( e^{-j\theta}y_i)},
\end{equation}
and where $\round{\cdot}$ rounds its argument to the nearest integer\footnote{The direction of rounding for half-integers is not important so long as it is consistent.  The authors have chosen to round up half-integers in their own implementation.}.  Note that $\hat{d}_i(\theta)$ does not depend on the amplitude $\rho$.  As defined, $\hat{u}_i(\theta)$ is not strictly inside the set of integers $\{0, 1, \dots, M-1\}$, but this is not of consequence, as we intend its value to be considered modulo $M$.  With this in mind let $\phi_i = \angle y_i$, then,
\[
\hat{u}_i(\theta) = \round{\frac{M}{2\pi}( \phi_i - \theta )}
\]
which is equivalent to the definition from~\eqref{eq:hatdfinxtheta} modulo $M$.

@@

\begin{lemma}
Let $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ so that $\hat{\theta}$ is the least squares estimator of the carrier phase and $\hat{\rho}$ is the least squares estimator of the amplitude.  Let 
\[
\lambda_i = \phi_i - \hat{\theta} - \frac{2\pi}{M}\round{\frac{M}{2\pi}(\phi_i - \hat{\theta})}.
\]
Then 
\[
\abs{\lambda_i} \leq \frac{1}{2M} - \frac{1}{2LM BLERG}
\]
for all $i \in P \cup D$.
\end{lemma}
\begin{proof}
\end{proof}

@@

\begin{figure*}[tp]
	\centering
		\includegraphics{code/plot-2.mps}
		\caption{Phase error}
		\label{fig:plotphase}
\end{figure*}

\begin{figure*}[tp]
	\centering
		\includegraphics{code/plot-3.mps}
		\caption{Complex gain error}
		\label{fig:plotcamp}
\end{figure*}

@@

\bibliography{bib}
